---
title: "LAB 11"
author: "STAT 131"
date: "November 21, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to the lab 11! In this lab, you will

- Do regression diagnosis on the datasets from the previous labs.

# Regression dianosis

## Red wine dataset

Let's first look at the red wine quality prediction dataset from lab 10.

Read data.

```{r}
wine<- read.csv("winequality-red.csv", sep = ";")
wine$quality <- wine$quality + rnorm(length(wine$quality))
```

Fit the model.
```{r}
wine.fit <- lm(quality~volatile.acidity+chlorides+free.sulfur.dioxide+total.sulfur.dioxide+pH+sulphates+alcohol, data=wine)
summary(wine.fit)
```

**Exercise 1**

(a) Do regression diagnostics using the `plot` function.

```{r}
# insert your code here to do regression diagnostics.

```

(b) Answer the following TRUE/FALSE questions based on the diagnostics plot. Uncomment your answer.

```{r}
### I. The plot indicates heteroscedasticity.
# TRUE
# FALSE
### II. There are non-linearity between the explantory variable and response variable.
# TRUE
# FALSE
### III. The normal assumtion holds for this model.
# TRUE
# FALSE
```

(c) Identify at least two outliers from the data. 

```{}

```
> I think the sample ??? and ??? are outliers.


## Diamond dataset

Now let us look at the diamond dataset from lab 9. 

Read the data.

```{r}
diamonds <- read.csv("diamonds.csv")
diamonds <- diamonds[sample(1:nrow(diamonds), 1000), ]
head(diamonds)
```

Fit a linear regression.

```{r}
diamond.fit <- lm(price ~ carat + cut + color + clarity + depth + table, data = diamonds)
summary(diamond.fit)
```

**Exercise 2**

(a) Do regression diagnostics using the `plot` function.

```{r}
# insert your code here to do regression diagnostics.

```

(b) Answer the following TRUE/FALSE questions based on the diagnostics plot. Uncomment your answer.

```{r}
### I. The plot indicates heteroscedasticity.
# TRUE
# FALSE
### II. There are non-linearity between the explantory variable and response variable.
# TRUE
# FALSE
### III. The normal assumtion holds for this model.
# TRUE
# FALSE
```

