---
title: "LAB 4"
author: "STAT 131"
date: "Sep 25, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in data.

```{r}
craigslist <- read.csv("craigslist.csv", 
                       header = TRUE, stringsAsFactors = FALSE)
one.bedrooms <- craigslist[craigslist$brs == 1, ]
```

### Simulation: when will the t-test fail

Following four functions generate four types of sample (You can ignore the details concerning how the samples are generated.)

- **generate.sample1**: Small sample size (5) and normal distributed samples.
- **generate.sample2**: Small sample size (5) and non-normal distributed samples.
- **generate.sample3**: Large sample size (100) and non-normal distributed samples.
- **generate.sample4**: Large sample size (100), normal distirbuted but sample not independent.

```{r}
generate.sample1 <- function(){
  sample <- rnorm(5)
  return(sample)
}

generate.sample2 <- function(){
  sample <- rgamma(5, 1.5, 2)
  return(sample)
}

generate.sample3 <- function(){
  sample <- rcauchy(100, 0, 1)
  return(sample)
}

generate.sample4 <- function(){
  sample <- arima.sim(n = 100, list(ar = c(1, -0.5), ma = c(0.3)))
  return(sample)
}
```

If you run the function, such as `generate.sample1`, it will generate the corresponding samples.

```{r}
samples <- generate.sample1()
samples
```


**Exercise 3.**

Please complete the function `simulation`. It accept the name of the above sample generation functions and return the estimated type I error. Please follow the following steps for completing the function `simulation`:

1. generate two samples from the function `FUN`.
2. calculate the p-value using t-test.
3. repeat 1 and 2 for N times.
4. calculate the chance of rejecting the null hypothesis $H_0:\mu_1 = \mu_2$ when the significant level is set to 0.05. (that is, the estimated Type I Error) HINT: The chance of rejecting the null can be estimated by ((The number of rejections) / N) = ((The number of p values less or equal than 0.05) / N).

```{r}
simulation <- function(FUN, N = 10000){
  # FUN: function name for sample generation
  # N: number of repetitions

  
}
```

\pagebreak

After complete your function above, you can run the following code:

```{r e31}
set.seed(123456) # please do not modify the seed
simulation(generate.sample1)
```

```{r e32}
set.seed(123456) # please do not modify the seed
simulation(generate.sample2)
```

```{r e33}
set.seed(123456) # please do not modify the seed
simulation(generate.sample3)
```

```{r e34}
set.seed(123456) # please do not modify the seed
simulation(generate.sample4)
```

Since the null hypothesis is true when we generate the data, the estimated Type I error should be close to 0.05 as we increase the number of repetitions. However, you may observe that t-test fails except for the first sample.  

From the simulation, which test do you think is better for the Craigslist dataset? why?

```{e35}
Write your non-coding answer here.

```

\pagebreak

# Bonferroni correction

**Exercise 4.**

(a) Perform t-test on the meam price difference for each pairwise city combinations. Save the p-values in vector `p.values`. For which city pairs we can not reject the null hypothesis that mean rent price is equal? (under level 0.05) Again, you can use `apply` or for loops, but `apply` is recommended.

```{r e4a1}
cities <- unique(craigslist$location) # get the vector of unique city names
cities.combn <- combn(cities, 2) # this will give you all the possible combinations of two cities
# insert code here save the number of postings for Berkeley as
# 'p.values'
p.values <- apply(cities.combn, 2, function(x){
  # The price vector for city 1 (x[1])
  # city1data <- 
  # The price vector for city 2 (x[2])
  # city2data <- 
  # return the p value of t test
  # p <- 
  # return(p)
})
```

```{r e4a2}
# You can not reject the null hypothesis that mean rent price for the following city pairs.
# (Subset the `cities.combn`.)
# save the your answer as
# 'not.reject'
not.reject = c()
```

\pagebreak

(b) Perform Bonferroni corrections on the p-values. For which city pairs we can not reject the null hypothesis that mean rent price is equal? (under level 0.05)

```{r e4b1}
# insert code here save you answer as
# 'p.values.adj'
p.values.adj = c()
```

```{r e4b2}
# You can not reject the null hypothesis that mean rent price for the following city pairs.
# (Subset the `cities.combn`.)
# (After Bonferroni corrections)
# insert code here save you answer as
# 'not.reject.adj'
not.reject.adj = c()
```

\pagebreak

```{r}
p.values
```

```{r}
not.reject
```

```{r}
p.values.adj
```

```{r}
not.reject.adj
```


