---
title: "Lab 10"
output:
  pdf_document: default
  html_notebook: default
---

```{r,echo=FALSE,warning=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
```


## *Exercise 1* 

We will continue using diamonds dataset

```{r}
diamonds <- read.csv("diamonds.csv")

diamonds$price = as.numeric(diamonds$price)

## Select Numeric Columns
diamonds = diamonds[, sapply(diamonds, class) == "numeric"]
head(diamonds)
```

```{r}
### Fitting a linear model
fit <- lm(price ~ ., data = diamonds)
summary(fit)

### Getting fitted values
fitted.vals = fit$fitted.values
```

### (a) 
Calculate the residuals, the residual sum of squares (RSS), and the total sum of squares (TSS) using the `fitted.value()`
```{r}
# Insert you code here
# RSS <-
# TSS <-
```

### (b) 
Calculate the R-square ($R^2$) using RSS and TSS. What is the interpretation of $R^2$?
```{r}
# Insert you code here, save your results as `Rsq`
# Rsq <-
```

### (c)
Fit another multivariate model (fit.restricted), but this time, drop `length.in.mm`, `width.of.mm` and `depth.in.mm`. Plot the residuals from this model against the variables added as covariates. 
```{r}
# fit.restricted <- lm(price ~)
#summary(fit.restricted)


```



# Regression dianosis

## Red wine dataset

Reload the data.
```{r}
wine<- read.csv("winequality-red.csv", sep = ";")
wine$quality <- wine$quality + rnorm(length(wine$quality))
```

Fit the model.
```{r}
wine.fit <- lm(quality~volatile.acidity+chlorides+free.sulfur.dioxide+total.sulfur.dioxide+pH+sulphates+alcohol, data=wine)
summary(wine.fit)
```


### Exercise 2

(a) Do regression diagnostics using the `plot` function.

```{r}
# insert your code here to do regression diagnostics.

```

(b) Answer the following TRUE/FALSE questions based on the diagnostics plot. Uncomment your answer.

```{r}
### I. The plot indicates heteroscedasticity.
# TRUE
# FALSE
### II. There are non-linearity between the explantory variable and response variable.
# TRUE
# FALSE
### III. The normal assumtion holds for this model.
# TRUE
# FALSE
```

(c) Identify at least two outliers from the data. 

```{r}

```

> I think the sample ??? and ??? are outliers.


## Multiple regression with continuous and categorical variables

### Exercise 3

(a) Fit a linear regression model with explanatory variable `carat`, `depth`, `table`, `clarity`, `color` and `cut`.

```{r}
diamonds <- read.csv("diamonds.csv")
head(diamonds,2)
```


```{r}
# Insert you code here, save your model as `fit.categorical`
# fit.categorical <- lm(price ~ .......


```

(b) Write the equation when 
	i. Clarity is VS2, color is H, and cut is Premium. Replace ??? with numerical values.
		$$\text{price} = ??? + ??? \cdot \text{carat} + ??? \cdot \text{depth} + ??? \cdot \text{table}$$
	ii. clarity is I1, color is D and cut is Fair. Replace ??? by numerical values.
		$$\text{price} = ??? + ??? \cdot \text{carat} + ??? \cdot \text{depth} + ??? \cdot \text{table}$$
		
		

## Diamond dataset 

We will include categorical variables in the following analysis.
Read the data.

```{r}
diamonds <- read.csv("diamonds.csv")
diamonds <- diamonds[sample(1:nrow(diamonds), 1000), ]
head(diamonds)
```

Fit a linear regression.

```{r}
diamond.fit <- lm(price ~ carat + cut + color + clarity + depth + table, data = diamonds)
summary(diamond.fit)
```

### Exercise 4

(a) Do regression diagnostics using the `plot` function.

```{r}
# insert your code here to do regression diagnostics.
```

(b) Answer the following TRUE/FALSE questions based on the diagnostics plot. Uncomment your answer.

```{r}
### I. The plot indicates heteroscedasticity.
# TRUE
# FALSE
### II. There are non-linearity between the explantory variable and response variable.
# TRUE
# FALSE
### III. The normal assumtion holds for this model.
# TRUE
# FALSE
```

